---
title: "Single cell data analysis"
author: "Fizza Irfan"
format:
  html:
    embed-resources: true
---


## DGE from Tijana

From her paper, fetch and unzip

``` bash
git clone https://github.com/tijanaran/NRAS-mutant_melanoma_Single-cell_RNA_Seq.git
for FILE in NRAS-mutant_melanoma_Single-cell_RNA_Seq/*.zip; do unzip "$FILE"; done
```


```{r}
#| label: setup
#| include: false

# Load necessary libraries
library(dplyr)
library(Seurat)
library(ggplot2)
library(readr)
library(tibble)
library(purrr)
library(fs)
```


Correspondences of experiments from Tijana `scRNA_TR.Rmd`,


```{r}
correspondences <- c(
  TISKM1_S1_DGE = "Untreated_SKMel30",
  TISKM2_S2_DGE = "Treated_24h_SKMel30",
  TISKM3_S3_DGE = "Treated_4days_SKMel30",
  TISKM4_S4_DGE = "Treated_long_SKMel30",
  TI21_S1_DGE = "Untreated_Meljuso",
  TI22_S2_DGE = "Treated_24h_Meljuso",
  TI23_S3_DGE = "Treated_4days_Meljuso",
  TI24_S4_DGE = "Treated_long_Meljuso",
  TI25_S1_DGE = "Untreated_IPC298",
  TI26_S2_DGE = "Treated_24h_IPC298",
  TI27_S3_DGE = "Treated_4days_IPC298",
  TI28_S4_DGE = "Treated_long_IPC298",
  TI41_S1_DGE = "Untreated_M20",
  TI42_S2_DGE = "Treated_24h_M20",
  TI43_S3_DGE = "Treated_4days_M20",
  TI44_S4_DGE = "Treated_long_M20"
)
```

```{r}
create_dgcmatrix <- function(path, max_cols = 1000) {
  id <- path_file(path) |> path_ext_remove() |> as.character()
  new_id <- correspondences[id]
  read_tsv(path, show_col_types = FALSE) |> 
    dplyr::select(seq_len(max_cols)) |> 
    column_to_rownames(var = "GENE") |> 
    as.matrix() |> 
    as("dgCMatrix") |> 
    CreateSeuratObject(project = new_id)
}
```

loop over the all files

```{r}
#| eval: false

fs::dir_ls("NRAS-mutant_melanoma_Single-cell_RNA_Seq/", regexp = "TI2[1-4].+_DGE\\.txt$") |> 
  map(create_dgcmatrix, max_cols = 1001, .progress = "Loading MelJUSO files") -> Meljuso

fs::dir_ls("NRAS-mutant_melanoma_Single-cell_RNA_Seq/", regexp = "TI2[5-8].+_DGE\\.txt$") |> 
  map(create_dgcmatrix, max_cols = 1001, .progress = "Loading IPC files") -> IPC298

fs::dir_ls("NRAS-mutant_melanoma_Single-cell_RNA_Seq/", glob = "*TISKM*_DGE.txt") |>
  map(create_dgcmatrix, max_cols = 1001,  .progress = "Loading SKMEL files") -> SKMel30
SKMel30

fs::dir_ls("NRAS-mutant_melanoma_Single-cell_RNA_Seq/", glob = "*TI4*_DGE.txt" ) |>
  map(create_dgcmatrix, max_cols = 1001, .progress = "Loading M20 files") -> M20
M20

saveRDS(Meljuso, "seurat_meljuso.rds")
saveRDS(IPC298, "seurat_ipc298.rds")
saveRDS(SKMel30, "seurat_skmel30.rds")
saveRDS(M20, "seurat_m20.rds")

```

Load saved data Seurat object per cell line

```{r}
Meljuso <- read_rds("seurat_meljuso.rds")
IPC298 <- read_rds("seurat_ipc298.rds")
SKMel30 <- read_rds("seurat_skmel30.rds")
M20 <- read_rds("seurat_m20.rds")
```



Current identifiers

```{r}
map_chr(Meljuso, \(x) Idents(x) |> levels())
map_chr(IPC298, \(x) Idents(x) |> levels())
map_chr(SKMel30, \(x) Idents(x) |> levels())
map_chr(M20, \(x) Idents(x) |> levels())
```

## Merge the Surat object without integration

```{r}
# Example to merge more than two Seurat objects
Meljuso_merge <- merge(x = Meljuso[[1]], y = Meljuso[-1])
IPC298_merge <- merge(x = IPC298[[1]], y = IPC298[-1])
SKMel30_merge <- merge(x = SKMel30[[1]], y = SKMel30[-1])
M20_merge <- merge(x = M20[[1]], y = M20[-1])
```



### defining function for filtering MT genes 

```{r}
# Define the function
filter_cell_lines <- function(cell_line) {
  cell_line[["percent.mt"]] <- PercentageFeatureSet(cell_line, pattern = "^MT-")
  cell_line_filtered <- subset(cell_line, subset = percent.mt < 15)
  
  return(cell_line_filtered)
}
```

## filtering MT genes for 4 cell lines
```{r}
Meljuso_filtered <- filter_cell_lines(Meljuso_merge)
IPC298_filtered <- filter_cell_lines(IPC298_merge)
SKMel30_filtered <- filter_cell_lines(SKMel30_merge)
M20_filtered <- filter_cell_lines(M20_merge)
```



###Visualize QC metrics as a violin plot

```{r}
#| warning: false
#| fig-width: 10
library(patchwork)

generate_plots <- function(seurat_obj) {
  vplot <- VlnPlot(seurat_obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
  plot1 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "percent.mt")
  plot2 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
  combined_plot <- (plot1 + plot2) / vplot
  return(combined_plot)
}
```

##Meljuso

```{r}
#| warning: false
#| fig-width: 10
Meljuso_plots <- generate_plots(Meljuso_filtered)
print(Meljuso_plots)
```

##IPC298
```{r}
#| warning: false
#| fig-width: 10
IPC298_plots <- generate_plots(IPC298_filtered)
print(IPC298_plots)
```

##SKMel30
```{r}
#| warning: false
#| fig-width: 10
SKMel30_plots <- generate_plots(SKMel30_filtered)
print(SKMel30_plots)
```

##M20
```{r}
#| warning: false
#| fig-width: 10
M20_plots <- generate_plots(M20_filtered)
print(M20_plots)
```
##Normalizing data
```{r}
Meljuso_normalized <- Meljuso_filtered %>%
    NormalizeData()
IPC298_normalized <-IPC298_filtered %>%
    NormalizeData()
SKMel30_normalized <-SKMel30_filtered %>%
    NormalizeData()
M20_normalized <-M20_filtered %>%
    NormalizeData()
```

## violin plot function
```{r}
plot_violin <- function(seurat_object, features, group_by = "orig.ident", pt_size = 0.1) {
  VlnPlot(seurat_object, features = features, layer = "data", group.by = group_by, pt.size = pt_size) +
    theme_minimal() +
    theme(plot.title = element_text(size = 15, face = "bold"))
}
```

## violin plots for normalized data

```{r}
Meljuso_violinplot1 <- plot_violin(Meljuso_normalized, features = c("BCL2", "PARP1"))
IPC298_violinplot1 <- plot_violin(IPC298_normalized, features = c("BCL2", "PARP1"))
SKMel30_violinplot1 <- plot_violin(SKMel30_normalized, features = c("BCL2", "PARP1"))
M20_violinplot1 <- plot_violin(M20_normalized, features = c("BCL2", "PARP1"))
```

```{r}
ls() |> stringr::str_subset("violinplot1$") |> 
  set_names(nm = \(x) stringr::str_remove(x, "_violinplot1$")) |> 
  map(\(x) {
    # convert characters back symbols and evaluate to get the object
    # then a list of 2, [[1]] is BCL2 and [[2]] is PARP
    # in ggplot, plot object are list, with the data.frame in the "data" element
    eval(as.name(x))[[1]][["data"]] |> as_tibble()
  }) |> 
  bind_rows(.id = "cells") |> 
  filter(stringr::str_detect(ident, "^Untreated")) |> 
  ggplot(aes(y = BCL2, x = ident, fill = ident)) +
  geom_violin() +
  geom_jitter(size = 0.5, alpha = 0.6) +
  theme_classic() +
  labs(
    title = "BCL-2 Expression in NRAS mutant cell lines",
    x = "Cell lines",
    y = "Normalized Expression"
  ) +
  theme(
    plot.title = element_text(size = 24, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 15, face = "bold"),
    axis.title.y = element_text(size = 15, face = "bold"),
    axis.text.x = element_text(size = 10, face = "bold", color = "black"),
    axis.text.y = element_text(size = 10, face = "bold", color = "black")
  ) #+
#scale_y_continuous(limits = c(0, 3))
```


#### Normalize the data and scale and PCA

```{r}
processing_seurat_object <- function(seurat_obj) {
  seurat_obj %>%
    NormalizeData() %>%
    FindVariableFeatures(selection.method = "vst", nfeatures = 2000) %>%
    ScaleData() %>%
    RunPCA()
}
```

## run PCA
```{r}
Meljuso_pca <- processing_seurat_object(Meljuso_filtered)
IPC298_pca <- processing_seurat_object(IPC298_filtered)
SKMel30_pca <- processing_seurat_object(SKMel30_filtered)
M20_pca <- processing_seurat_object(M20_filtered)
```


### visualize PCA results

```{r}
#print(SKMel30_pca[["pca"]], dims = 1:5, nfeatures = 5)
#VizDimLoadings(SKMel30_pca, dims = 1:2, reduction = "pca")
#DimPlot(SKMel30_pca, reduction = "pca")
```


### Perform UMAP

- Find neighbors and clusters

```{r}
UMAP_function <- function(seurat_object, dims = 1:10, resolution = 0.5) {
  seurat_object <- RunUMAP(seurat_object, dims = dims)
  seurat_object <- FindNeighbors(seurat_object, dims = dims)
  seurat_object <- FindClusters(seurat_object, resolution = resolution)
  
  return(seurat_object)
}
```

### apply UMAP on cell lines
```{r}
Meljuso_umap <- UMAP_function(Meljuso_pca)
IPC298_umap <- UMAP_function(IPC298_pca)
SKMel30_umap <- UMAP_function(SKMel30_pca)
M20_umap <- UMAP_function(M20_pca)
```

### Visualize BCL2 expression on UMAP

```{r}
#plot_feature_umap <- function(seurat_object, feature, group.by = "ident") {
  #FeaturePlot(seurat_object, features = feature, reduction = "umap") +
    #ggtitle(paste(feature, "Expression on UMAP")) +
    #theme_minimal() +
    #theme(plot.title = element_text(size = 15, face = "bold"))
#}
```



```{r}
plot_feature_umap <- function(seurat_object, feature, group.by = "orig.ident") {
  # Plot the UMAP with cell population highlighting
  dim_plot <- DimPlot(seurat_object, reduction = "umap", group.by = group.by) +
    ggtitle("Conditions") +
    theme_minimal() +
    theme(plot.title = element_text(size = 15, face = "bold"))

  # Overlay feature expression on the UMAP
  feature_plot <- FeaturePlot(seurat_object, features = feature, reduction = "umap") +
    ggtitle(paste(feature, "Expression on UMAP")) +
    theme_minimal() +
    theme(plot.title = element_text(size = 15, face = "bold"))
  
  # Combine both plots
  combined_plot <- cowplot::plot_grid(dim_plot, feature_plot, ncol = 2)
  
  return(combined_plot)
}


```



## Apply to cell lines
```{r}
Meljuso_BCL2_umap <- plot_feature_umap(Meljuso_umap, "BCL2")
IPC298_BCL2_umap <- plot_feature_umap(IPC298_umap, "BCL2")
SKMel30_BCL2_umap <- plot_feature_umap(SKMel30_umap, "BCL2")
M20_BCL2_umap <- plot_feature_umap(M20_umap, "BCL2")

print(Meljuso_BCL2_umap)
print(IPC298_BCL2_umap)
print(SKMel30_BCL2_umap)
print(M20_BCL2_umap)
```
## violin plot function

```{r}
plot_violin <- function(seurat_object, features, group_by = "orig.ident", pt_size = 0.1) {
  VlnPlot(seurat_object, features = features, layer = "data", group.by = group_by, pt.size = pt_size) +
    theme_minimal() +
    theme(plot.title = element_text(size = 15, face = "bold"))
}
```

##violin plot for meljuso
```{r}
Meljuso_violinplot <- plot_violin(Meljuso_umap, features = c("BCL2", "PARP1"))
Meljuso_violinplot
```

## violin plot for IPC298
```{r}
IPC298_violinplot <- plot_violin(IPC298_umap, features = c("BCL2", "PARP1"))
IPC298_violinplot
```

## violin plot for SKMel30
```{r}
SKMel30_violinplot <- plot_violin(SKMel30_umap, features = c("BCL2", "PARP1"))
SKMel30_violinplot
```

## violin plot for M20
```{r}
M20_violinplot <- plot_violin(M20_umap, features = c("BCL2", "PARP1"))
M20_violinplot
```


Extract expression of BCL2 from violin plot

```{r}
ls() |> stringr::str_subset("violinplot$") |> 
  set_names(nm = \(x) stringr::str_remove(x, "_violinplot$")) |> 
  map(\(x) {
    # convert characters back symbols and evaluate to get the object
    # then a list of 2, [[1]] is BCL2 and [[2]] is PARP
    # in ggplot, plot object are list, with the data.frame in the "data" element
    eval(as.name(x))[[1]][["data"]] |> as_tibble()
  }) |> 
  bind_rows(.id = "cells") |> 
  filter(stringr::str_detect(ident, "^Untreated")) |> 
  ggplot(aes(y = BCL2, x = ident, fill = ident)) +
  geom_violin() +
  geom_jitter(size = 0.5, alpha = 0.6) +
  theme_classic() +
  labs(
    title = "BCL-2 Expression in NRAS mutant cell lines",
    x = "Cell lines",
    y = "Normalized Expression after UMAP"
  ) +
  theme(
    plot.title = element_text(size = 24, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 15, face = "bold"),
    axis.title.y = element_text(size = 15, face = "bold"),
    axis.text.x = element_text(size = 10, face = "bold", color = "black"),
    axis.text.y = element_text(size = 10, face = "bold", color = "black")
  ) #+
#scale_y_continuous(limits = c(0, 3))
```



```{r}
ls() |> stringr::str_subset("violinplot$") |> 
  set_names(nm = \(x) stringr::str_remove(x, "_violinplot$")) |> 
  map(\(x) {
    # convert characters back symbols and evaluate to get the object
    # then a list of 2, [[1]] is BCL2 and [[2]] is PARP
    # in ggplot, plot object are list, with the data.frame in the "data" element
    eval(as.name(x))[[2]][["data"]] |> as_tibble()
  }) |> 
  bind_rows(.id = "cells") |> 
  filter(stringr::str_detect(ident, "^Untreated")) |> 
  ggplot(aes(y = PARP1, x = ident, fill = ident)) +
  geom_violin(trim = FALSE, alpha = 0.3) +
  geom_boxplot(width = 0.1, outlier.shape = NA, color = "black") +
  geom_jitter(size = 0.3, alpha = 0.3) +
  theme_classic() +
  labs(
    title = "PARP1 Expression Across Different Conditions",
    x = "Cell lines",
    y = "PARP1 Expression"
  ) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 15),
    axis.title.y = element_text(size = 15)
  )
```




